Black-swan events in animal populations

### 2016-07-05 text/ms_pnas.tex
Copyediting of pnas ms

\affil[a]{Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada, e-mail:
sandrsn@uw.edu}
\affil[b]{School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA}
\affil[c]{School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Black swans---statistically improbable events
with profound consequences---happen
more often than expected in financial, social, and natural systems. Our work
demonstrates the rare but systematic presence of black-swan events in animal populations
around the world (mostly birds, mammals, insects, and fishes). These events are
predominantly downward, implying that unexpected population crashes occur more
frequently than increases. Black-swan events are not driven by life history (e.g. lifespan),
but by external causes such as extreme winters and disease. Ignoring the presence of
downward black swans may severely underestimate extinction risk in animal populations,
particularly under a changing climate where such extreme events are expected to increase
in frequency and magnitude.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn\@uw.edu}

@@

\begin{abstract}
Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
\end{abstract}

@@

We found evidence that black-swan dynamics are infrequent but highly
influential: in our dataset they were detected most
frequently for birds (7\%), mammals (5\%), and
insects (3\%) but almost never in fishes (Fig.~2). Black-swan
events were taxonomically widespread, occurring in 38\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---four of
eight orders with the highest median probability of heavy tails were insect
orders (Fig.~3a).

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.3\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (b)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

Given the prevalence of downwards events, we refit our heavy-tailed models to
measure the degree of skewness \(\gamma\) of the process noise using a skew-t
distribution (Fig.~4a), and used these models to make near-term risk forecasts.
Aggregated across populations with strong evidence of heavy tails (median \(\nu
< 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating
strong evidence of downwardly skewed process noise (Fig.~4b,
Fig.~S\figskewnu). In contrast, populations that did not have heavy
tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
(Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these
heavy-tailed populations forward five years revealed that assuming standard
normal process noise underestimated risk (99\% lower credible interval of
abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
Fig.~S\figskewprojections).

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy-tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full ess %%% generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \cite{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
(e.g.~the $t$ distribution compared to a mixture
of two normal distributions \cite{ward2007}).
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution.
Furthermore, if the degrees of freedom parameter ($\nu$)
is fixed,
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 4$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 4$ allows for an event
three standard deviations below the mean to happen
once every 50 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
The Gompertz model represents population growth as a linear function in
log space (here and throughout we use log
to refer to the natural logarithm).
If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Studen}t\mhyphen t(\nu, 0, \sigma)
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

We chose the $t$ distribution for a number of reasons.
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations is not present.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \cite{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes
that could give rise to extreme events in population dynamics,
such as extreme climate events
\cite{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses
would be important for diagnosing the underlying causes
of extreme events and to develop predictive models,
it would be impossible to tease out each of these
underlying causes from population trend data alone.
Instead we seek a simple flexible phenomenological model
that can be applied across the widest possible
range of population time series to provide a
comparative overview of the prevalence
of black-swan events.

@@

For the Gompertz model, our weakly-informative priors (Fig.~S\figpriors) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by Ref.~\cite{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \cite{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

We chose a weakly informative exponential prior on \(\nu\) with rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. These thresholds are in
excess of the minimums recommended by Ref.~\cite{gelman2006a} of
\(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
confidence intervals. In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance itself (Supporting Information).

In addition to alternative population models, we investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$; Fig.~S\figpriors, Supporting Information). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \cite{ferrari2004}. The model was as follows:
\begin{align}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \text{for } i = 1, \dots, 609\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to match the
way they are visually represented in Fig.~S\figcorrelates\ and to make the
relationship approximately linear on the logit-transformed response scale. All
input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c,
gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more
probability density around expected parameter values while allowing for a
higher probability density far into the tails, thereby allowing the data to
dominate the posterior more strongly if it disagrees with the prior. We fit our
models with 5,000 total iterations per chain, 2500 warm-up iterations, four
chains, and discarding every second sample. We checked for chain convergence
visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
against a similar model with the glmmADMB \textsf{R} package \cite{glmmadmb}
without uncertainty around the predictors.

>>>>

\affil[a]{Earth to Ocean Research Group, Department of Biological
Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada}
\affil[b]{School of Aquatic and Fishery Sciences, Box 355020,
University of Washington, Seattle, WA 98195, USA}
\affil[c]{School of Resource and Environmental Management, Simon
Fraser University, Burnaby, BC, V5A 1S6, Canada}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Black swans---statistically improbable events
with profound consequences---happen
more often than expected in financial, social, and natural systems. Our work
demonstrates the rare but systematic presence of black-swan events in animal populations
around the world (mostly birds, mammals, insects, and fishes). These events are
predominantly downward, implying that unexpected population crashes occur more
frequently than increases. Black-swan events are not driven by life history (e.g.\ lifespan),
but by external causes such as extreme winters and disease. Ignoring the presence of
downward black swans may severely underestimate extinction risk in animal populations,
particularly under a changing climate where such extreme events are expected to increase
in frequency and magnitude.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{S.C.A. and T.A.B. conceived the project;
S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the
data and wrote the paper with input from all authors.}
\authordeclaration{Please declare any conflict of interest here.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: sandrsn@uw.edu}

@@

\begin{abstract}
Black swans are improbable events that nonetheless
occur---often with profound consequences. 
% While physical extremes, such as
% monsoons and heat waves, are widely studied and increasing in magnitude and frequency, 
Such events drive important transitions in social systems (e.g.\ banking
collapses) and physical systems (e.g.\ earthquakes)
yet it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. 
Here we estimate the prevalence and
direction of black-swan events (heavy-tailed process noise)
in \NPops\ animal populations
after accounting for population dynamics (productivity, 
density dependence, and typical stochasticity).
We find strong evidence for 
black-swan events in \(\sim\)\overallBasePerc \% of populations.
These events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
and are not explained by any life-history covariates, but
tend to be driven by external perturbations such as climate, severe winters,
predators, parasites, or the combined effects of multiple factors.
Black-swan events manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and ignoring heavy-tailed process noise leads to 
% a \crashUnderMedian-fold (50\% interval \crashUnderRange) 
an underestimate in the magnitude of population crashes.
We suggest modelers consider heavy-tailed downward-skewed 
probability distributions, 
such as the skewed Student-$t$ used here, 
when making forecasts of population abundance.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
\end{abstract}

@@

We found that black-swan dynamics are infrequent but highly
influential: in our dataset they were detected most
frequently for birds (7\%), mammals (5\%), and
insects (3\%) but almost never in fishes (Fig.~2). Black-swan
events were taxonomically widespread, occurring in 38\% of taxonomic orders.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a hierarchical model, we found stronger evidence
for black swans in insect populations than these statistics suggest---four of
eight orders with the highest median probability of heavy tails were insect
orders (Fig.~3a).

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.26\textwidth]{../analysis/order-posteriors-covariates.pdf}
\caption{Standardized probabilities and covariates of heavy-tailed
dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (b)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.
}
\label{fig:3}
\end{figure}

@@

Given the prevalence of downwards events, we refit our heavy-tailed models to
measure the degree of skewness \(\gamma\) of the process noise using a skew-$t$
distribution (Fig.~4a), and used these models to make near-term risk forecasts.
Aggregated across populations with strong evidence of heavy tails (median \(\nu
< 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating
strong evidence of downwardly skewed process noise (Fig.~4b,
Fig.~S\figskewnu). In contrast, populations that did not have heavy
tails (median \(\nu \geq 70\)) had little evidence of skewed process noise
(Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these
heavy-tailed populations forward five years revealed that assuming standard
normal process noise underestimated risk (99\% lower credible interval of
abundance) by \crashUnderRange-fold (interquartile range; Fig.~4c,
Fig.~S\figskewprojections).

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy-tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full ess %%% generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \cite{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
(e.g.~the $t$ distribution compared to a mixture
of two normal distributions \cite{ward2007}).
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution.
Furthermore, if the degrees of freedom parameter ($\nu$)
is fixed,
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 5$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 5$ allows for an event
three standard deviations below the mean to happen
% 1/(1-pt(3, 5))
once every 66 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

@@

We fit heavy-tailed Gompertz population dynamics models to data from the GPDD.
The Gompertz model represents population growth as a linear function in
log space (here and throughout we use log
to refer to the natural logarithm).
If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then
\begin{align}
x_t &= \lambda + bx_{t - 1} + \epsilon_t\\
  \epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
The parameter \(\lambda\) represents the expected growth rate if \(N_{t}
= 1\). The model is density independent if \(b = 1\), maximally density
dependent if \(b = 0\), and inversely density dependent if \(b < 0\). Usually,
the process noise \(\epsilon_{t}\) is modelled as normally distributed, but we
allow it to be drawn from a Student-$t$ distribution with scale parameter
\(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to
which the process deviations have heavy tails and are therefore evidence of
black-swan events (Fig.~1A, B).

We chose the $t$ distribution for a number of reasons.
Pragmatically, the $t$ distribution requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations is absent.
Furthermore, the $t$ distribution arises from a mixture of normal distributions
with the same mean and different variances \cite{gelman2014},
which mimics the underlying basis for black-swan events.
There are a wide range of mechanistic processes
that could give rise to extreme events in population dynamics,
such as extreme climate events
\cite{meehl2004, katz2005, ipcc2012}
or sudden changes in predation from other species.
While evaluating these various hypotheses
would be important for diagnosing the underlying causes
of extreme events and to develop predictive models,
it would be impossible to tease out each of these
underlying causes from population trend data alone.
Instead we seek a simple flexible phenomenological model
that can be applied across the widest possible
range of population time series to provide a
comparative overview of the prevalence
of black-swan events.

@@

For the Gompertz model,
we chose weakly informative priors for 
all parameters (Fig.~S\figpriors, Supporting Information).
For the degrees of freedom parameter $\nu$, 
we chose an exponential prior 
with a rate parameter
of \(0.01\) truncated at values above two---a slightly less informative prior
than suggested by Ref.~\cite{fernandez1998}. We truncated the
distribution at \(2\), since at \(\nu < 2\) the variance of the $t$ distribution
is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\)
but constrains the sampling sufficiently to avoid \(\nu\) drifting towards infinity. In
any case, for \(\nu > 20\) the $t$ distribution is almost indistinguishable from
the normal distribution (Fig.~1). Based on the shape of the $t$ distribution, we
chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the
probability of heavy-tailed (black-swan) dynamics. When categorizing a population as heavy
tailed, we applied a 0.5 probability threshold. In the scenario where the data
are uninformative about heavy tails (e.g.~Fig.~1E, H), the posterior for
\(\nu\) will approximately match the prior (prior median \(= 71\), mean \(=
102\)) and a metric of Pr\((\nu < 10) > 0.5\) will not flag the
population as heavy tailed.

We fit our models with Stan 2.6.0 \cite{stan-manual2015, hoffman2014}, and R
3.2.0 \cite{r2015}. We began with four chains and 2,000 iterations, discarding
the first 1,000 as warm up (i.e.~4,000 final samples). If \(\hat{R}\) (the
potential scale reduction factor---a measure of chain convergence) was greater
than \(1.05\) for any parameter, or if the minimum effective sample size,
\(n_\mathrm{eff}\) (a measure of the effective number of uncorrelated samples),
for any parameter was less than \(200\), we doubled both the total iterations
and warm up period and sampled from the model again. 
% These thresholds are in
% excess of the minimums recommended by Ref.~\cite{gelman2006a} of
% \(\hat{R} < 1.1\) and \(n_\mathrm{eff} > 100\) for reliable point estimates and
% confidence intervals.
In the majority of cases our minimum thresholds were
greatly exceeded. We continued this procedure up to 8,000 iterations (16,000
total samples) by which all chains had sufficiently converged.
The No-U-Turn Hamiltonian Markov chain Monte Carlo sampler in Stan generally
requires fewer iterations to obtain equivalent effective sample sizes than the
Gibbs or Metropolis--Hastings algorithms \cite{stan-manual2015}.

\subsection{Alternative models and simulation testing}
We fit alternative population models to test if four key phenomena
systematically changed our conclusions. Autocorrelation has been suggested as
a reason for increased observed variability of abundance time series through
time, which could create apparent heavy tails \citep{inchausti2002};
therefore, we fit a model that included serial correlation in the residuals.
Additionally, previous work has modelled abundance or growth rates without
accounting for density dependence \citep{halley2002,segura2013}; therefore, we
fit a simpler model in which we assumed density independence. Third,
observation error could bias parameter estimates \citep{knape2012} or mask our
ability to detect heavy tails \citep{ward2007}; therefore, we fit a model
where we allowed for a fixed quantity of observation error ($0.2$ standard
deviations on a log scale). Finally, the Gompertz model assumes that
population growth rate declines linearly with log abundance. Therefore, we
also fit an alternative model, the Ricker-logistic model, which assumes that
population growth rate declines linearly with abundance (Supporting Information).

We investigated the sensitivity
of our results to weaker and stronger priors (exponential rate parameter $=
0.005, 0.02$; Fig.~S\figpriors, Supporting Information). Furthermore, we used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Information).

\subsection{Covariates of heavy-tailed dynamics} We fit a hierarchical beta
regression model to the predicted probability of heavy tails, Pr(\(\nu < 10\)),
to investigate potential covariates of heavy-tailed dynamics. We obtained
maximum lifespan and body-size data from Ref.~\cite{brook2006a}.
Beta regression is useful when response data are on a continuous scale
between zero and one \cite{ferrari2004}. The model was as follows:
\begin{align*}
\mathrm{Pr}(\nu_i < 10) &\sim \mathrm{Beta}(A_i, B_i)\\
\mu_i &= \mathrm{logit}^{-1}(\alpha
  + \alpha^\mathrm{class}_{j[i]}
  + \alpha^\mathrm{order}_{k[i]}
  + \alpha^\mathrm{species}_{l[i]}
  + \bm{X_i} \bm{\beta}),
  \: \\
  &\quad \, \text{for } i = 1, \dots, 609\\
A_i &= \phi_\mathrm{disp} \mu_i\\
B_i &= \phi_\mathrm{disp} (1 - \mu_i)\\
\alpha^\mathrm{class}_j &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{class}}),
  \: \text{for } j = 1, \dots, 6\\
\alpha^\mathrm{order}_k &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{order}}),
  \: \text{for } k = 1, \dots, 38\\
\alpha^\mathrm{species}_l &\sim
  \mathrm{Normal}(0, \sigma^2_{\alpha \; \mathrm{species}}),
  \: \text{for } l = 1, \dots, 301,
\end{align*}
where \(A\) and \(B\) represent the beta distribution shape parameters;
\(\mu_i\) represents the predicted value for population \(i\), class \(j\),
order \(k\), and species \(l\); \(\phi_\mathrm{disp}\) represents the
dispersion parameter; and \(\bm{X_i}\) represents a vector of predictors (such as
lifespan) for population \(i\) with associated \(\bm{\beta}\) coefficients. The
intercepts are allowed to vary from the overall intercept \(\alpha\) by
taxonomic class (\(\alpha^\mathrm{class}_j\)), taxonomic order
(\(\alpha^\mathrm{order}_k\)), and species (\(\alpha^\mathrm{species}_l\)) with
standard deviations \(\sigma_{\alpha \; \mathrm{class}}\), \(\sigma_{\alpha \;
\mathrm{order}}\), and \(\sigma_{\alpha \; \mathrm{species}}\). Where possible,
we also allowed for error distributions around the predictors by incorporating
the standard deviation of the posterior samples for the Gompertz parameters
\(\lambda\), \(b\), and \(\log \sigma\) around the mean point value as normal
distributions (not shown in the above equation).

We log transformed \(\sigma\), time-series length, and lifespan to 
to make the
relationship approximately linear on the logit-transformed response scale;
this matches the way the predictors are visually presented in 
Fig.~S\figcorrelates. 
All input variables were standardized by subtracting their mean and dividing by two
standard deviations to make their coefficients comparable in
magnitude \cite{gelman2008c}. We excluded body length as a covariate because it
was highly correlated with lifespan, and lifespan exhibited more overlap across
taxonomy than body length. Lifespan is also more directly related to time and
potential mechanisms driving black-swan dynamics.

We incorporated weakly informative priors into our model: \(\mathrm{Cauchy}(0,
10)\) on the global intercept \(\alpha\), \(\mathrm{half\mhyphen Cauchy}(0,
2.5)\) on all standard deviation parameters, \(\mathrm{half\mhyphen Cauchy}(0,
10)\) on the dispersion parameter \(\phi_\mathrm{disp}\), and
\(\mathrm{Cauchy}(0, 2.5)\) on all other parameters \cite{gelman2006c,
gelman2008d}. 
% Compared to normal priors, the Cauchy priors concentrate more
% probability density around expected parameter values while allowing for a
% higher probability density far into the tails, thereby allowing the data to
% dominate the posterior more strongly if it disagrees with the prior. 
We fit our
models with 5,000 total iterations per chain, 2500 warm-up iterations, four
chains, and discarding every second sample. We checked for chain convergence
visually and with the same criteria as before (\(\widehat{R} < 1.05\) and
\(n_\mathrm{eff} >200\) for all parameters). Furthermore, we checked our model
against a similar model with the glmmADMB \textsf{R} package \cite{glmmadmb}
without uncertainty around the predictors.
